Computing exact restricted isometry constants via mixed integer semidefinite programming

### 2014-12-18 RIP-MISDP-Paper/Paper_v1.tex
remove \\

\section{Introduction}
Finding the sparsest solution of an underdetermined system of linear equations is a problem that has been the subject of much recent research in compressed sensing. Given a matrix $A \in \R^{m \times n}$ with $m < n$ and 
right-hand side $b \in \R^m$, we are interested in a solution of 
\begin{align}\label{l0min}
 \text{min} \quad & \|x\|_0 \\
 \text{s.t.} \quad & Ax = b, \nonumber
\end{align}
where $\|x\|_0$ stands for the cardinality and thus the number of nonzero components of a vector $x$. As it is $\mathcal{NP}$-hard to solve (\ref{l0min}), see for example \cite{GareyJohnson}, we would like to solve the problem
\begin{align}\label{l1min}
 \text{min} \quad & \|x\|_0 \\
 \text{s.t.} \quad & Ax = b, \nonumber
\end{align}
instead, as this can be cast as an LP. Results in compressed sensing show, that there are many conditions which imply that the problems (\ref{l0min}) and (\ref{l1min}) are indeed equivalent. First results about this have been 
established in \cite{DH01}, and have later been followed up for example by \cite{CRT05}, \cite{Zha08} or \cite{BCT11}. One such condition is the restricted isometry property
\begin{equation}\label{RIPintro}
 (1-\delta_k)\|x\|_2^2 \leq \|Ax\|_2^2 \leq (1+\delta_k)\|x\|_2^2,
\end{equation}
that was first introduced in \cite{CT05}, if it holds for all $k$-sparse vectors $x$ for a sufficiently small restricted isometry constant $\delta_k$, then problems (\ref{l0min}) and (\ref{l1min}) have the same optimal solution. 
Even tough it was shown in \cite{BDDW08}, that this condition holds for multiple classes of random matrices with probability exponentially going to one, in \cite{PT14} it was shown, that computing the optimal restricted isometry 
constants in (\ref{RIPintro}) for a given matrix $A$ is again an $\mathcal{NP}$-hard task. \\
In the past SDP-Relaxations were proposed in \cite{Asp07} and \cite{Asp08} to compute upper bounds on the values of the restricted isometry 
constant, which could then be used to verify the equivalance between (\ref{l0min}) and (\ref{l1min}), but there could also be cases, where the results of those relaxations weren't sufficient, but the real constants would have been. 
In this paper we will construct a mixed integer semidefinite program to compute the exact constants in (\ref{RIPintro}), to be able to verify in all cases if those sufficient conditions for the equivalence hold for a given matrix. 
The main contribution of this paper is theorem \ref{Rk1thm}, which shows that the optimal solution of that MISDP indeed allows us to compute the exact restricted isometry constant.\\
In the following we will start by introducing the restricted isometry property and corresponding results about the equivalence of (\ref{l0min}) and (\ref{l1min}). In the next section we will present the SDP-relaxations in \cite{Asp07} 
and \cite{Asp08}.Afterwards we will construct a mixed integer semidefinite programme and show that it produces exact restricted isometry constants before presenting numerical results and comparisons with the SDP-Relaxations in 
\cite{Asp07} and \cite{Asp08}.

@@

Again this result can be generalized to the case of imperfect data. One advantage of Defintion \ref{asymRIPdef} and Theorem \ref{asymRIPtheorem} is, that they are indifferent under scaling of the system of linear equations. If a system
$Ax=b$ is scaled to $\lambda Ax = \lambda b$, then the condition of Theorem \ref{RIPsqrt2} might no longer be valid, while in Definition \ref{asymRIPdef} we can use $\tilde{\alpha_k} = \lambda \alpha_k, \tilde{\beta_k} = \lambda \beta_k$
and therefore $\tilde{\gamma_k} = \lambda^2 \alpha_k^2 / \lambda^2 \beta_k^2 = \gamma_k$. \\
In the following we will try to compute the optimal constants $\alpha_k$ and $\beta_k$ in (\ref{asymRIP}). These can then be used to compute either the restricted isometry ratio $\gamma_k := \beta_k^2 / \alpha_k^2$ or the restricted
isometry constant $\delta_k = \text{max} \{ 1 - \alpha_k^2, \beta_k^2 - 1 \}$.

@@

\begin{thm}\label{Rk1thm}
 For every $A \in \R^{m \times n}$ and $k \geq 1$ there exists an optimal solution $X^*, z^*$ to (\ref{MISDP}) such that $\textbf{Rk}(X^*) = 1$. 
\end{thm}
\begin{proof}
 Let $\hat{X}, \hat{z}$ be an optimal solution to (\ref{MISDP}) (because of $\textbf{Tr}(X) = 1$ the problem is bounded and $X_{11} = 1, X_{ij} = 0 \ \forall i,j \neq 1, z_1 = 1, z_i = 0 \ \forall i \neq 0$ is always feasible)
 with $\sum_{i=1}^n z_i = \ell \leq k$. Define
 \begin{align}\label{Proj}
 \text{min} \quad & \textbf{Tr}(\tilde{A}\T\tilde{A}X) \nonumber \\
 \text{s.t.} \quad & \textbf{Tr}(X) = 1 \tag{Proj} \\
 & X \succeq 0 , X \in S_\ell, \nonumber
 \end{align}
where $\tilde{A} = (A_{i_1}, A_{i_2}, ..., A_{i_\ell})$ consists of those columns $\{{i_1}, {i_2}, ..., {i_\ell}\} =: T$ of $A$ for which $\hat{z}_{i_k} = 1$. \\
Now let $\tilde{X}$ be the matrix consisting of those rows and columns 
$i_k$ of $\hat{X}$ such that $\hat{z}_{i_k} = 1$. Then $\hat{X}$ is feasible for (\ref{Proj}) with the same objective value as the optimal solution $\hat{X}$ of (\ref{MISDP}), as we only removed zero rows and columns of $\hat{X}$ 
and $(A\T A)\hat{X}$, which do not influence the trace or the positive semidefiniteness. \\
On the other hand we can lift any optimal solution $\tilde{X}$ of (\ref{Proj}) to a feasible point for (\ref{MISDP}) with identical objective value by setting $\bar{z}_i$ to one if and only if $i \in T$
and $\bar{X}_{i_j i_k} = \tilde{X}_{jk}$ if and only if $i_j, i_k \in T$ and $\bar{X}_{ij}=0$ otherwise. \\
So the optimal objective values of (\ref{Proj}) and (\ref{MISDP}) agree and with this construction we also get that 
$\textbf{Rk}(\bar{X}) = \textbf{Rk}(\tilde{X})$, so it suffices to show that (\ref{Proj}) always allows for a solution of rank one.
Let $\breve{X}$ be an extreme point of (\ref{Proj}) (as the set of optimal points of (\ref{Proj}) is a non-empty, convex, compact subset of $\R^{\ell(\ell+1)/2}$ such a point exists as a consequence of corollary 18.5.1 in
\cite{Roc70}) with corresponding face $F=\{\breve{X}\}$. Then using theorem \ref{patakiLemma} on (\ref{Proj}) with $\tilde{m} = 1$ LP-constraint gives
\begin{equation*}
 \frac{1}{2}\textbf{Rk}(\breve{X})\cdot(\textbf{Rk}(\breve{X})+1) \leq \tilde{m} + \textbf{dim}(F) = 1 + 0 = 1.
\end{equation*}
As $\breve{X}=0$ is infeasible for $\textbf{Tr}(X)=1$, we must have $\textbf{Rk}(\breve{X}) = 1$, which can then be lifted to an optimal solution of (\ref{MISDP}) with rank one.
\end{proof}

>>>>

\section{Introduction}
Finding the sparsest solution of an underdetermined system of linear equations is a problem that has been the subject of much recent research in compressed sensing. Given a matrix $A \in \R^{m \times n}$ with $m < n$ and 
right-hand side $b \in \R^m$, we are interested in a solution of 
\begin{align}\label{l0min}
 \text{min} \quad & \|x\|_0 \\
 \text{s.t.} \quad & Ax = b, \nonumber
\end{align}
where $\|x\|_0$ stands for the cardinality and thus the number of nonzero components of a vector $x$. As it is $\mathcal{NP}$-hard to solve (\ref{l0min}), see for example \cite{GareyJohnson}, we would like to solve the problem
\begin{align}\label{l1min}
 \text{min} \quad & \|x\|_0 \\
 \text{s.t.} \quad & Ax = b, \nonumber
\end{align}
instead, as this can be cast as an LP. Results in compressed sensing show, that there are many conditions which imply that the problems (\ref{l0min}) and (\ref{l1min}) are indeed equivalent. First results about this have been 
established in \cite{DH01}, and have later been followed up for example by \cite{CRT05}, \cite{Zha08} or \cite{BCT11}. One such condition is the restricted isometry property
\begin{equation}\label{RIPintro}
 (1-\delta_k)\|x\|_2^2 \leq \|Ax\|_2^2 \leq (1+\delta_k)\|x\|_2^2,
\end{equation}
that was first introduced in \cite{CT05}, if it holds for all $k$-sparse vectors $x$ for a sufficiently small restricted isometry constant $\delta_k$, then problems (\ref{l0min}) and (\ref{l1min}) have the same optimal solution. 
Even tough it was shown in \cite{BDDW08}, that this condition holds for multiple classes of random matrices with probability exponentially going to one, in \cite{PT14} it was shown, that computing the optimal restricted isometry 
constants in (\ref{RIPintro}) for a given matrix $A$ is again an
$\mathcal{NP}$-hard task.

In the past SDP-Relaxations were proposed in \cite{Asp07} and \cite{Asp08} to compute upper bounds on the values of the restricted isometry 
constant, which could then be used to verify the equivalance between (\ref{l0min}) and (\ref{l1min}), but there could also be cases, where the results of those relaxations weren't sufficient, but the real constants would have been. 
In this paper we will construct a mixed integer semidefinite program to compute the exact constants in (\ref{RIPintro}), to be able to verify in all cases if those sufficient conditions for the equivalence hold for a given matrix. 
The main contribution of this paper is theorem \ref{Rk1thm}, which shows
that the optimal solution of that MISDP indeed allows us to compute the
exact restricted isometry constant.

In the following we will start by introducing the restricted isometry property and corresponding results about the equivalence of (\ref{l0min}) and (\ref{l1min}). In the next section we will present the SDP-relaxations in \cite{Asp07} 
and \cite{Asp08}.Afterwards we will construct a mixed integer semidefinite programme and show that it produces exact restricted isometry constants before presenting numerical results and comparisons with the SDP-Relaxations in 
\cite{Asp07} and \cite{Asp08}.

@@

Again this result can be generalized to the case of imperfect data. One advantage of Defintion \ref{asymRIPdef} and Theorem \ref{asymRIPtheorem} is, that they are indifferent under scaling of the system of linear equations. If a system
$Ax=b$ is scaled to $\lambda Ax = \lambda b$, then the condition of Theorem \ref{RIPsqrt2} might no longer be valid, while in Definition \ref{asymRIPdef} we can use $\tilde{\alpha_k} = \lambda \alpha_k, \tilde{\beta_k} = \lambda \beta_k$
and therefore $\tilde{\gamma_k} = \lambda^2 \alpha_k^2 / \lambda^2
\beta_k^2 = \gamma_k$.

In the following we will try to compute the optimal constants $\alpha_k$ and $\beta_k$ in (\ref{asymRIP}). These can then be used to compute either the restricted isometry ratio $\gamma_k := \beta_k^2 / \alpha_k^2$ or the restricted
isometry constant $\delta_k = \text{max} \{ 1 - \alpha_k^2, \beta_k^2 - 1 \}$.

@@

\begin{thm}\label{Rk1thm}
 For every $A \in \R^{m \times n}$ and $k \geq 1$ there exists an optimal solution $X^*, z^*$ to (\ref{MISDP}) such that $\textbf{Rk}(X^*) = 1$. 
\end{thm}
\begin{proof}
 Let $\hat{X}, \hat{z}$ be an optimal solution to (\ref{MISDP}) (because of $\textbf{Tr}(X) = 1$ the problem is bounded and $X_{11} = 1, X_{ij} = 0 \ \forall i,j \neq 1, z_1 = 1, z_i = 0 \ \forall i \neq 0$ is always feasible)
 with $\sum_{i=1}^n z_i = \ell \leq k$. Define
 \begin{align}\label{Proj}
 \text{min} \quad & \textbf{Tr}(\tilde{A}\T\tilde{A}X) \nonumber \\
 \text{s.t.} \quad & \textbf{Tr}(X) = 1 \tag{Proj} \\
 & X \succeq 0 , X \in S_\ell, \nonumber
 \end{align}
where $\tilde{A} = (A_{i_1}, A_{i_2}, ..., A_{i_\ell})$ consists of those
columns $\{{i_1}, {i_2}, ..., {i_\ell}\} =: T$ of $A$ for which
$\hat{z}_{i_k} = 1$.

Now let $\tilde{X}$ be the matrix consisting of those rows and columns 
$i_k$ of $\hat{X}$ such that $\hat{z}_{i_k} = 1$. Then $\hat{X}$ is feasible for (\ref{Proj}) with the same objective value as the optimal solution $\hat{X}$ of (\ref{MISDP}), as we only removed zero rows and columns of $\hat{X}$ 
and $(A\T A)\hat{X}$, which do not influence the trace or the positive
semidefiniteness.

On the other hand we can lift any optimal solution $\tilde{X}$ of (\ref{Proj}) to a feasible point for (\ref{MISDP}) with identical objective value by setting $\bar{z}_i$ to one if and only if $i \in T$
and $\bar{X}_{i_j i_k} = \tilde{X}_{jk}$ if and only if $i_j, i_k \in T$
and $\bar{X}_{ij}=0$ otherwise.

So the optimal objective values of (\ref{Proj}) and (\ref{MISDP}) agree and with this construction we also get that 
$\textbf{Rk}(\bar{X}) = \textbf{Rk}(\tilde{X})$, so it suffices to show that (\ref{Proj}) always allows for a solution of rank one.
Let $\breve{X}$ be an extreme point of (\ref{Proj}) (as the set of optimal points of (\ref{Proj}) is a non-empty, convex, compact subset of $\R^{\ell(\ell+1)/2}$ such a point exists as a consequence of corollary 18.5.1 in
\cite{Roc70}) with corresponding face $F=\{\breve{X}\}$. Then using theorem \ref{patakiLemma} on (\ref{Proj}) with $\tilde{m} = 1$ LP-constraint gives
\begin{equation*}
 \frac{1}{2}\textbf{Rk}(\breve{X})\cdot(\textbf{Rk}(\breve{X})+1) \leq \tilde{m} + \textbf{dim}(F) = 1 + 0 = 1.
\end{equation*}
As $\breve{X}=0$ is infeasible for $\textbf{Tr}(X)=1$, we must have $\textbf{Rk}(\breve{X}) = 1$, which can then be lifted to an optimal solution of (\ref{MISDP}) with rank one.
\end{proof}
